---
title: "Integrating Spatial Data on Malaria Vectors into Malaria Risk Mapping Approaches"
author: "Ernest Moyo"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

This document implements a Bayesian mechanistic modeling approach using the Ross-McDonald equations integrated within the **EpiWave FOI Model** framework. It leverages spatial covariates to refine high-resolution malaria risk mapping as part of Ernest Moyo’s PhD research at NM-AIST and Vector Atlas.

# Load Required Libraries

```{r}
library(greta)
library(greta.dynamics)
library(greta.gp)
library(ggplot2)
```

# Ross-McDonald Model Integration

## Differential Equations

The Ross-McDonald model equations are:

```{r}
dx_dt <- function(x, z, m, a, b, r) {
  m * a * b * z * (1 - x) - r * x
}

dz_dt <- function(x, z, a, c, g) {
  a * c * x * (1 - z) - g * z
}
```

## State Iteration

```{r}
iterate_state <- function(state, iter, m, a, b, c, r, g) {
  x <- state[1]
  z <- state[2]

  state_new <- c(
    x + dx_dt(x, z, m, a, b, r),
    z + dz_dt(x, z, a, c, g)
  )
  state_new
}
```

## Bayesian Integrated Simulation Function

```{r}
simulate_integrated_malaria_model <- function(n_days = 365, n_sims = 100, X_covariate) {

  # Priors
  alpha <- normal(0, 1)
  beta <- normal(1, 1, truncation = c(0, Inf))

  log_m <- alpha + beta * X_covariate
  m <- exp(log_m)

  a <- normal(0.2, 0.05, truncation = c(0,1))
  b <- normal(0.6, 0.05, truncation = c(0,1))
  c <- normal(0.9, 0.05, truncation = c(0,1))
  inv_r <- normal(14, 1, truncation = c(0, Inf))
  inv_g <- normal(7, 1, truncation = c(0, Inf))

  r <- 1/inv_r
  g <- 1/inv_g

  initial_state <- c(0.01, 0)

  solution <- iterate_dynamic_function(
    transition_function = iterate_state,
    initial_state = initial_state,
    niter = n_days,
    tol = 0,
    m = m,
    a = a,
    b = b,
    c = c,
    r = r,
    g = g,
    parameter_is_time_varying = "m"
  )

  x <- solution$all_states[1, ]
  z <- solution$all_states[2, ]

  log_S <- log(m * a * b * t(z))
  epsilon <- gp(matrix(1:n_days), kernel = rbf(lengthscale = 30, variance = 1))

  log_xi <- log_S + epsilon
  xi <- exp(log_xi)

  M_human <- normal(1000, 100)
  I_lt <- M_human * xi
  gamma_lt <- normal(0.1, 0.01)

  C_lt <- poisson(I_lt * gamma_lt)

  sims <- calculate(x, z, xi, C_lt, nsim = n_sims)

  par(mfrow=c(4,1))

  plot(sims$x[1,1,], type='n', ylim=c(0,1), main="Humans Infected (x)", ylab="Proportion infected", xlab="Day")
  for(i in 1:n_sims) lines(sims$x[i,1,])

  plot(sims$z[1,1,], type='n', ylim=c(0,1), main="Mosquitoes Infected (z)", ylab="Proportion infected", xlab="Day")
  for(i in 1:n_sims) lines(sims$z[i,1,])

  plot(sims$xi[1,1,], type='n', main="Force of Infection (xi)", ylab="FOI", xlab="Day")
  for(i in 1:n_sims) lines(sims$xi[i,1,])

  plot(sims$C_lt[1,1,], type='n', main="Simulated Malaria Cases (C_lt)", ylab="Cases", xlab="Day")
  for(i in 1:n_sims) lines(sims$C_lt[i,1,])

  return(list(
    humans_infected = sims$x,
    mosquitoes_infected = sims$z,
    force_of_infection = sims$xi,
    simulated_cases = sims$C_lt
  ))
}
```

# Example Simulation

Below is an example simulation using synthetic sinusoidal covariate data to mimic seasonal variation:

```{r}
X_covariate <- sin(2 * pi * (1:365) / 365)
results <- simulate_integrated_malaria_model(n_days=365, n_sims=10, X_covariate=X_covariate)
```

# Conclusion

This integrated Bayesian approach using the Ross-McDonald equations forms a foundational component of high-resolution malaria risk mapping. The model supports targeted intervention and surveillance optimization, aligning closely with the specific objectives of the PhD project.

------------------------------------------------------------------------

*For inquiries:* **Ernest Moyo**\
Email: `ernestmoyo35@gmail.com` \| \`[moyoe\@nm-aist.ac.tz](mailto:moyoe@nm-aist.ac.tz){.email}
